Reasonably Programmable Literal Notation

### 2018-06-19 icfp18/syntax-icfp18.tex
edits to 2.1.3-2.1.6 and new diagram figure

%% Cyrus packages
\usepackage{microtype}
\usepackage{todonotes}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{wasysym}
\definecolor{light-gray}{gray}{0.95}

@@

\subsubsection{Context Dependence}
\label{sec:context-dependence}
The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on. In this case, Lines 5-7 of Fig.~\ref{fig:regex-tlm-def} specify that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally. In  general, the dependency can be an arbitrary module path, e.g. \li{module List = Core.Std.List}. The \li{Pervasives} module is implicitly opened in each expansion. All other bindings, both at the TLM definition site and the TLM application site, are not internally available to the expansion.

@@

From the client programmer's perspective, the benefit of an application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope. This is a common problem with various unhygienic approaches, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.

The benefits of making the macro definition site dependencies explicit, rather than implicitly allowing expansions to access all bindings at the definition site as in many existing macro systems, are more subtle. One technical benefit is that this exposes the TLM's dependencies in the signature of the module where the TLM is defined. This is useful for build tools that extract dependencies by source code analysis. Moreover, if an ``internal'' module is being used by expansions, then this will be manifest in the signature and can be corrected if this was not intended. 
% (Similar concerns about unintended capture drives recent work on explicit closures for distributed computing \cite{DBLP:conf/ecoop/MillerHO14}.) 
Implicit access to the  definition site would require the implementation to carefully ``smuggle'' internal values to each application site, thereby skirting the abstraction discipline of ML's module system \cite{culpepper2005syntactic}.% (and making the implementation substantially more complex!)\todo{I think there is some work on this...}  %This, in turn, ensures that the expansion does not depend on bindings inaccessible outside the module, i.e. the implementation does not need to ``smuggle'' references to these internal definitions to each application site. Consequently, the abstraction discipline of the ML module system is maintained.

Another benefit of this explicit approach is that renamings can be propagated without needing to modify AST encodings in parser definitions. For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to propagate to the module paths in the TLM definition in the usual manner as long as the names of the internal module variables remain the same. (The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). This allows us to encode expansions using ordinary OCaml values (see Sec. \ref{sec:provider-perspective}).

Enforcing this strong context independence condition is technically subtle because TLM parsers need to be able to implement splicing, i.e. they need to be able to parse  terms out of the literal body for placement in the expansion. Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately constrain application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site. For example, consider the final expansion of our example from Line 2 of Fig.~\ref{fig:first-tlm-example}:
\begin{lstlisting}[numbers=none]
  Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))
\end{lstlisting}
In this term, both \li{Regex} and \li{DNA} are free module variables. There is nothing to distinguish references to \li{DNA} that arose from a spliced sub-expression parsed out of the literal body from those that would indicate that the context independence condition has been violated. 
%\footnote{An exhaustive search for "DNA" in the literal body would not be  because final expansions are abstract syntax trees. There is not a unique mapping back to concrete syntax, e.g. because whitespace can be inserted arbitrarily.}

Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason. These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}. Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro for rewriting. %From the macro expander's perspective, it has appeared by magic''.

To address this problem, TLM parsers are not tasked with generating the final expansion directly. Instead, the parser generates a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body. For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as follows:
\begin{lstlisting}[numbers=none]
  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
\end{lstlisting}
Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}. We return to the type annotation on the splice reference, \li{Regex.t}, in Sec. \ref{sec:segment-typing} below. The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency, so all is well.

@@

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the supplement, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\vspace{-3px}
\label{fig:nesting-example}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service. For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g. for Emacs, Vim, and so on), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicit, this information can be reported even when there is a parse error or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

Segment types are somewhat analogous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body. This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}), as intended.% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

To address this problem, splicing is guaranteed to be capture-avoiding. The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion via capture-avoiding substitution, which automatically alpha-varies the internal bindings of the proto-expansion as necessary. There is no need for TLM providers to manually deploy a mechanism that generates fresh variables (as in, e.g., Racket's reader macros \cite{Flatt:2012:CLR:2063176.2063195}, further discussed in Sec. \ref{sec:existing-approaches}). For example, the final expansion of the example above is alpha-equivalent to the following:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... application site temporary ... */; 
  let tmp_fresh = /* ... expansion-internal temporary ... */;
  Html.H1Element(tmp_fresh, f(tmp));
\end{lstlisting}
%Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp'}. The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.

@@

Although this strict capture avoidance discipline implies that TLMs cannot intentionally introduce  bindings directly into spliced expressions, this does not imply that values cannot flow from the expansion into a spliced expression. It simply means that when this is intended, the segment type must be a function type, which serves to make this interface explicit. Reason's concise lambda notation, \li{(x) => e}, decreases the syntactic cost of this approach. For example, we cannot define list comprehension notation like the following:
\begin{lstlisting}[numbers=none,deletekeywords={in}]
  $listcomp `(x + 1SURL | x in EURLloSURL .. EURLhi)` /* NO! cannot reason abstractly */
\end{lstlisting}
However, the following is permitted, because \li{x} is bound by the spliced lambda expression and the corresponding segment type makes the type of the interface explicit:
\begin{lstlisting}[numbers=none]
  $listcomp `((x) => x + 1 SURL|EURL lo SURL..EURL hi)`  /* OK! */
\end{lstlisting}

Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation cannot possibly be performing ``magic'' on the type and binding structure of the language. We say more about the future prospect of a  mechanism designed specifically for shorthand binding forms, e.g. Haskell-style \li{do} notation \cite{jones2003haskell}, in Sec. \ref{sec:discussion}.% For now, the existing infix operator support in Reason and OCaml suffices for working with monadic values.

@@

@@

The second of these two premises {typechecks} the proto-expression or proto-pattern, and simultaneously generates a corresponding core language expression or pattern, from which the final expansion is constructed as described above. This step is specified by the following judgements:% are types and expanded expressions, respectively.
\[\begin{array}{ll}
\cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
% \cvalidR{\Delta}{\Gamma}{\escenev}{\crv}{r}{\tau}{\tau'} & \text{$\crv$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
\end{array}\]

@@

Notice that proto-rules, $\crv$, involve expanded patterns, $p$, not proto-patterns, $\cpv$, because proto-rules appear in proto-expressions, which are generated by expression TLMs. Proto-patterns arise only from pattern TLMs. There is no variable proto-pattern form, for the reasons described in Sec. \ref{sec:sptsms}.

\paragraph{Splice References} 
\begin{figure}
\begin{minipage}{\textwidth}
\small
$\arraycolsep=3pt\begin{array}{llcl}
\mathsf{PrTyp} & \ctau & ::= & t ~\vert~
\aceparr{\ctau}{\ctau} ~\vert~ \cdots ~\vert~
% \aceall{t}{\ctau} ~\vert~
% \acerec{t}{\ctau} ~\vert~
% \aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
% & & \vert & 
% \acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~ 
\acesplicedt{m}{n}\\
\mathsf{PrExp} & \ce & ::= & x ~\vert~
% \aceasc{\ctau}{\ce} ~\vert~
% \aceletsyn{x}{\ce}{\ce} ~\vert~
\acelam{\ctau}{x}{\ce} ~\vert~
\aceap{\ce}{\ce} ~\vert~
\cdots ~\vert~ 
% \acetlam{t}{\ce} ~\vert~
% \acetap{\ce}{\ctau} ~\vert~
% \acefold{\ce} \\
% & & \vert & \acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}} ~\vert~
% \acein{\ell}{\ce} ~\vert~
\acematchwith{n}{\ce}{\seqschemaX{\crv}} ~\vert~
\acesplicede{m}{n}{\ctau}\\
\mathsf{PrRule} & \crv & ::= & \acematchrule{p}{\ce}\\
\mathsf{PrPat} & \cpv & ::= & \acewildp ~\vert~
\cdots ~\vert~
% \acefoldp{p} ~\vert~
% \acetplp{\labelset}{\mapschema{\cpv}{i}{\labelset}} ~\vert~
% \aceinjp{\ell}{\cpv} ~\vert~
\acesplicedp{m}{n}{\ctau} 
\end{array}$
\end{minipage}
\caption[Syntax of proto-types and proto-expressions]{Syntax of proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
\label{fig:U-candidate-terms}
\end{figure}
The only interesting forms in Fig.~\ref{fig:U-candidate-terms} are the references to spliced types, expressions and patterns. Let us consider the rule for references to spliced expressions:

@@

>>>>

%% Cyrus packages
\usepackage{microtype}
\usepackage{todonotes}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage[all]{xy}
\usepackage{wasysym}
\definecolor{light-gray}{gray}{0.95}

@@

\subsubsection{Context Dependence}
\label{sec:context-dependence}

The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on. In this case, Line 6 of Fig.~\ref{fig:regex-tlm-def} specifies that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally. In  general, the dependency can be an arbitrary module path, e.g. \li{module List = Core.Std.List}. The \li{Pervasives} module is implicitly opened in each expansion. All other bindings, both at the TLM application site and the TLM definition site, are not internally available to the expansion.

@@

From the client programmer's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.

The benefits of making the macro definition site dependencies explicit, rather than implicitly allowing expansions to access all bindings at the definition site as in many existing macro systems, arise from the fact that this exposes the TLM's dependencies in the signature of the module where the TLM is defined. This is useful for build tools that extract dependencies by source code analysis. Moreover, if an ``internal'' module is being used by expansions, then this will be manifest in the signature and can be corrected if this was unintended. 
% (Similar concerns about unintended capture drives recent work on explicit closures for distributed computing \cite{DBLP:conf/ecoop/MillerHO14}.) 
Implicit access to the  definition site would require the implementation to carefully ``smuggle out'' otherwise internal values to each application site, thereby skirting the abstraction discipline of ML's module system \cite{culpepper2005syntactic}.% (and making the implementation substantially more complex!)\todo{I think there is some work on this...}  %This, in turn, ensures that the expansion does not depend on bindings inaccessible outside the module, i.e. the implementation does not need to ``smuggle'' references to these internal definitions to each application site. Consequently, the abstraction discipline of the ML module system is maintained.

\begin{figure}
\vspace{-5px}
\begin{displaymath}
\begin{array}{l}
  \text{\li{$regex `(GC$(DNA.any_base)GC)`}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The applied TLM lexes and parses the literal body to generate the \textbf{proto-expansion}.}} \\ ~
  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The system performs proto-expansion validation and generates the \textbf{final expansion}.}} \\ ~
  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))}}
\end{array}
\end{displaymath}
\caption{Expansion of TLM applications occurs in two steps.}
\vspace{-5px}
\label{fig:expansion}
\end{figure}

Enforcing this strong context independence condition is technically subtle because TLM parsers need to be able to implement splicing, i.e. they need to be able to parse  terms out of the literal body for placement in the expansion. Na\"ively checking that only the explicitly named dependencies are free in the expansion would inappropriately constrain application site spliced expressions, which should certainly not be prevented from referring to variables in scope at the application site. For example, consider the  bottom of Fig.~\ref{fig:expansion}, which shows the final expansion of the example from Line 2 of Fig.~\ref{fig:first-tlm-example}. 
% \begin{lstlisting}[numbers=none]
%   Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))
% \end{lstlisting}
In this term, both \li{Regex} and \li{DNA} are free module variables. There is nothing to distinguish references to \li{DNA} that arose from a spliced sub-expression parsed out of the literal body from those that would indicate that the context independence condition has been violated. 
%\footnote{An exhaustive search for "DNA" in the literal body would not be  because final expansions are abstract syntax trees. There is not a unique mapping back to concrete syntax, e.g. because whitespace can be inserted arbitrarily.}

To address this problem, TLM parsers do not generate the final expansion directly, but rather a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body. For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as shown in the middle of Fig.~\ref{fig:expansion}. 
% \begin{lstlisting}[numbers=none]
%   Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
% \end{lstlisting}
Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}. We return to the type annotation on the splice reference, \li{Regex.t}, in Sec. \ref{sec:segment-typing} below. The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.

@@

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the supplement, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\label{fig:nesting-example}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g. for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

Segment types are somewhat analogous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body. This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

To address this problem, splicing is guaranteed to be capture-avoiding. The final expansion is generated by recursively expanding each spliced expression and then inserting it into the final expansion via capture-avoiding substitution, which automatically alpha-varies the internal bindings of the proto-expansion as necessary. There is no need for TLM providers to manually deploy a mechanism that generates fresh variables (as in, e.g., Racket's reader macros \cite{Flatt:2012:CLR:2063176.2063195}, further discussed in Sec. \ref{sec:existing-approaches}). For example, the final expansion of the example above is alpha-equivalent to the following:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... application site temporary ... */; 
  let tmp_fresh = /* ... expansion-internal temporary ... */;
  Html.H1Element(tmp_fresh, f(tmp));
\end{lstlisting}
Notice that the expansion-internal binding of \li{tmp} has been alpha-varied to \li{tmp_fresh}. The reference to \li{tmp} in the spliced expression then refers, as intended, to the application site binding.

@@

Although this strict capture avoidance discipline implies that TLMs cannot intentionally introduce  bindings directly into spliced expressions, this does not imply that values cannot flow from the expansion into a spliced expression. It simply means that when this is intended, the segment type must be a function type, which serves to make this interface explicit. Reason's concise lambda notation, \li{(x) => e}, decreases the syntactic cost of this approach. For example, we cannot define list comprehension notation like the following because the binding site of \li{x} is not clear:
\begin{lstlisting}[numbers=none,deletekeywords={in}]
  $listcomp `(x + 1SURL | x in EURLloSURL .. EURLhi)` /* NO! cannot reason abstractly */
\end{lstlisting}
However, the following is permitted, because \li{x} is bound by the spliced lambda expression and the corresponding segment type makes the type of the interface explicit:
\begin{lstlisting}[numbers=none]
  $listcomp `((x) => x + 1 SURL|EURL lo SURL..EURL hi)`  /* OK! */
\end{lstlisting}

Our contention is that small syntactic costs like these are more than justified by the peace of mind of knowing that unfamiliar literal notation cannot possibly be performing ``magic'' with the type and binding structure. We say more about the future prospect of a  mechanism designed specifically for shorthand binding forms, e.g. Haskell-style \li{do} notation \cite{jones2003haskell}, in Sec. \ref{sec:discussion}.% For now, the existing infix operator support in Reason and OCaml suffices for working with monadic values.

@@

 \todo{integrate me}{}Another benefit of this explicit approach is that renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions. For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the TLM definition, as long as the name of the internal module variable in the dependencies list remains \li{Regex}. 
%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec. \ref{sec:provider-perspective}).

@@

The second of these two premises {typechecks} the proto-expression or proto-pattern, and simultaneously generates a corresponding core language expression or pattern, from which the final expansion is constructed as described above. This step is specified by the following judgements:% are types and expanded expressions, respectively.
\[\begin{array}{ll}
\cvalidT{\Delta}{\tscenev}{\ctau}{\tau} & \text{$\ctau$ has well-formed expansion $\tau$}\\
\cvalidE{\Delta}{\Gamma}{\escenev}{\ce}{e}{\tau} & \text{$\ce$ has expansion $e$ of type $\tau$}\\
% \cvalidR{\Delta}{\Gamma}{\escenev}{\crulev}{r}{\tau}{\tau'} & \text{$\crulev$ has expansion $r$ taking values of type $\tau$ to values of type $\tau'$}\\
\cvalidP{\upctx}{\pscenev}{\cpv}{p}{\tau} & \text{$\cpv$ has expansion $p$ matching $\tau$}
\end{array}\]

@@

Notice that proto-rules, $\crulev$, involve expanded patterns, $p$, not proto-patterns, $\cpv$, because proto-rules appear in proto-expressions, which are generated by expression TLMs. Proto-patterns arise only from pattern TLMs. There is no variable proto-pattern form, for the reasons described in Sec. \ref{sec:sptsms}.

\paragraph{Splice References} 
\begin{figure}
\begin{minipage}{\textwidth}
\small
$\arraycolsep=3pt\begin{array}{llcl}
\mathsf{PrTyp} & \ctau & ::= & t ~\vert~
\aceparr{\ctau}{\ctau} ~\vert~ \cdots ~\vert~
% \aceall{t}{\ctau} ~\vert~
% \acerec{t}{\ctau} ~\vert~
% \aceprod{\labelset}{\mapschema{\ctau}{i}{\labelset}} \\
% & & \vert & 
% \acesum{\labelset}{\mapschema{\ctau}{i}{\labelset}} ~\vert~ 
\acesplicedt{m}{n}\\
\mathsf{PrExp} & \ce & ::= & x ~\vert~
% \aceasc{\ctau}{\ce} ~\vert~
% \aceletsyn{x}{\ce}{\ce} ~\vert~
\acelam{\ctau}{x}{\ce} ~\vert~
\aceap{\ce}{\ce} ~\vert~
\cdots ~\vert~ 
% \acetlam{t}{\ce} ~\vert~
% \acetap{\ce}{\ctau} ~\vert~
% \acefold{\ce} \\
% & & \vert & \acetpl{\labelset}{\mapschema{\ce}{i}{\labelset}} ~\vert~
% \acein{\ell}{\ce} ~\vert~
\acematchwith{n}{\ce}{\seqschemaX{\crulev}} ~\vert~
\acesplicede{m}{n}{\ctau}\\
\mathsf{PrRule} & \crulev & ::= & \acematchrule{p}{\ce}\\
\mathsf{PrPat} & \cpv & ::= & \acewildp ~\vert~
\cdots ~\vert~
% \acefoldp{p} ~\vert~
% \acetplp{\labelset}{\mapschema{\cpv}{i}{\labelset}} ~\vert~
% \aceinjp{\ell}{\cpv} ~\vert~
\acesplicedp{m}{n}{\ctau} 
\end{array}$
\end{minipage}
\caption[Syntax of proto-types and proto-expressions]{Syntax of proto-expansions. Proto-expansion terms are ABTs identified up to alpha-equivalence.}
\label{fig:U-candidate-terms}
\end{figure}
The only interesting forms in Fig.~\ref{fig:U-candidate-terms} are the references to spliced types, expressions and patterns. Let us consider the rule for references to spliced expressions:

@@

\todo{integrate me}{}Hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for composite literal notation at other types for exactly this reason. These systems would find that the appearance of \li{DNA.any_base}  violates their context independence condition (which is one aspect of {hygiene}), because \li{DNA.any_base} is not, from the perspective of the context-free syntax, an argument to the macro nor even a sub-term of an argument for which a tree path can be assigned \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,gorn1965explicit}. Instead, it arises as the result of performing a complex operation---parsing---on an arbitrary sub-sequence of the string literal passed into the macro. (It is  possible for a macro to selectively opt out of the hygiene discipline---see Sec.~\ref{sec:existing-approaches} for more on why relying on this sort of ``escape hatch'' is an unreasonable solution.) %From the macro expander's perspective, it has appeared by magic''.
